---
title: "Exam One Grades"
author: "JBrandon Duck-Mayr"
date: "`r quack::american_date_format(Sys.Date())`"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}[1]{\setlength{\itemsep}{#1}}
    - \newcommand{\setskip}[1]{\setlength{\parskip}{#1}}
    - \renewcommand{\tightlist}{\setsep{3pt}\setskip{3pt}}
output:
    quack::presentation:
        toc: false
        incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
quack::attach_common_packages()
```

## Grade Distribution

```{r}
grades = readr::read_csv("curve.csv") %>% 
    mutate(Exam_Percent = Exam1 * 5) %>%
    rename(Curve = power_curve) %>% 
    select(Student, Exam_Percent, Curve) %>% 
    pivot_longer(cols = !Student, names_to = "Curved", values_to = "Grade") %>% 
    mutate(Curve = ifelse(Curved == "Curve", "Yes", "No"))
pal1 = c(Yes = "#d55e00", No = "#0072b2")
pal2 = c(Yes = "#d55e0030", No = "#0072b230")
mean1 = mean(grades$Grade[grades$Curve == "No"])
mean2 = mean(grades$Grade[grades$Curve == "Yes"])
median1 = median(grades$Grade[grades$Curve == "No"])
median2 = median(grades$Grade[grades$Curve == "Yes"])
ggplot(data = grades, mapping = aes(x = Grade, color = Curve, fill = Curve)) +
    geom_density() +
    geom_vline(xintercept = mean1, color = "#0072b2") +
    geom_vline(xintercept = median1, color = "#0072b2", linetype = "dashed") +
    geom_vline(xintercept = mean2, color = "#d55e00") +
    geom_vline(xintercept = median2, color = "#d55e00", linetype = "dashed") +
    scale_color_manual(name = "Curved?", values = pal1) +
    scale_fill_manual(name = "Curved?", values = pal2) +
    coord_cartesian(ylim = c(0, 0.05), expand = FALSE) +
    theme_bw()
```

## Grade Distribution

```{r}
tbl_dat = grades %>% 
    group_by(Curve) %>% 
    summarize(
        `1st quartile` = quantile(Grade, probs = 0.25),
        Mean = mean(Grade),
        Median = median(Grade),
        `75th Percentile` =quantile(Grade, probs = 0.75)
    ) %>% 
    arrange(desc(Curve))
kableExtra::kable(tbl_dat, digits = 0, booktabs = TRUE) %>%
    kableExtra::kable_styling(position = "center")
```

## Hard Questions

- Some questions were harder than others
- I'll go over the most missed questions
- The essays were good, minimal feedback needed

## Hard Questions: Constitutional Change (Process)

How do we change the United States Constitution?

. . .

+ Formal amendment process
+ Precedent
+ Importance: Difficulties in amendment help shape institutions and policy

## Hard Questions: Establishing the Judiciary

True or False? Article III of the Constitution establishes the U.S. Supreme Court and the lower federal courts.

. . .

+ It establishes the Supreme Court and lets Congress fill in the rest
+ Importance: Congress can add/remove courts, change # of justices, etc

## Hard Questions: Change by the Court

In no more than one or a few words, identify a constitutional development we discussed of each of these types:
Supreme Court interpretation (not spurred by other actors)

. . .

+ The Warren Court's expansion of liberty is a good answer
+ Importance: Most major constitutional developments are not the Court acting alone, but some are
