---
title: "Presidential Selection and Changes to the Bureaucracy"
author: "JBrandon Duck-Mayr"
date: "`r quack::american_date_format(Sys.Date())`"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}[1]{\setlength{\itemsep}{#1}}
    - \newcommand{\setskip}[1]{\setlength{\parskip}{#1}}
    - \renewcommand{\tightlist}{\setsep{1pt}\setskip{2pt}}
    - \usepackage{subcaption}
    # - "\\newcommand{\\fullpic}[2]{\\includegraphics[height=#1\\linewidth]{#2}}"
output:
    quack::presentation:
        toc: false
        incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Selecting the President

. . .

- The President must be (Article II \S\ 1 cl. 5)
  + a "natural born citizen", or one when the Constitution was ratified
  + 35 years old
  + Resided in the US for 14 years
- Presidents are elected for four year terms (Article II \S\ 1 cl. 1)
  + Unless impeached and removed from office (Article II \S\ 4)
- Presidents are elected by an "Electoral College"

## The Electoral College (original)

We originally had the following scheme for selecting the president:

\begin{block}{Summary of the Original Article II \S 1, cl. 3}
\begin{itemize}
    \item Each elector casts two ballots for president
    \item If only one person has a vote from a majority of electors, that person is selected as president
    \item If more than one person has such a number of votes, the House of Representatives chooses the president (from that set of people)
    \item If no one has such a number of votes, the House chooses from among the top five
    \item Any voting by the House is by state delegation (each state gets one vote)
    \item Then the person with the most votes becomes vice president; the Senate breaks ties
\end{itemize}
\end{block}

## Federalists vs. Democratic-Republicans (Again)

\begin{figure}
    \centering
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=0.3\linewidth,height=0.45\linewidth]{slides/unit-3/washington.jpg}
        \includegraphics[width=0.3\linewidth,height=0.45\linewidth]{slides/unit-3/adams.jpg}
        \includegraphics[width=0.3\linewidth,height=0.45\linewidth]{slides/unit-3/hamilton.jpg}
        \caption*{Federalists}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=0.3\linewidth,height=0.45\linewidth]{slides/unit-3/jefferson.jpg}
        \includegraphics[width=0.3\linewidth,height=0.45\linewidth]{slides/unit-3/burr.jpg}
        \includegraphics[width=0.3\linewidth,height=0.45\linewidth]{slides/unit-3/madison.jpg}
        \caption*{Democratic-Republicans}
    \end{subfigure}
\end{figure}

## The Election of 1796

- Adams got the most electoral votes
- Jefferson got the second most
- Under the original system, that meant Adams, a Federalist became president, and Jefferson, one of the founders of the Democratic-Republican party, became vice president

## The Election of 1800

- Jefferson and Burr ran as Democratic-Republican candidates; all Democratic-Republicans knew the intention was for Jefferson to be president and Burr to be vice president
- However, they tied in the electoral college; so, it went to the House
- The House vote failed to produce a winner for **35 ballots**
- Eventually, through back-channel dealings, Jefferson was able to gain the support of 10 (of 18) states' delegations
- Interestingly, Hamilton was among those pushing for Jefferson to win

## A Tragic Epilogue

\begin{center}
    \includegraphics[height = 0.5\textheight]{slides/unit-3/duel.jpg}
\end{center}

## The Twelth Amendment

Now we have the following scheme:

\begin{block}{Summary of Amendment XII}
\begin{itemize}
    \item Each elector casts a ballot for president and a ballot for vice president
    \item If a person has a majority of votes for president, that person is selected as president
    \item If no one does, the House chooses from among the top three
    \item Any voting by the House is by state delegation (each state gets one vote)
    \item If a person has a majority of votes for vice president, that person is selected as vice president
    \item If not, the Senate chooses from among the top two
\end{itemize}
\end{block}

## Bush v. Gore: Facts

. . .

\begin{center}
    \includegraphics[height = 0.9\textheight]{slides/unit-3/recount-paper}
\end{center}

## Bush v. Gore: Facts

\begin{center}
    \includegraphics[height = 0.85\textheight]{slides/unit-3/florida2000ballot.png}
\end{center}

## Bush v. Gore: Facts

\begin{center}
    \includegraphics[height = 0.85\textheight]{slides/unit-3/chads.jpg}
\end{center}

## Bush v. Gore: Facts

\begin{center}
    \includegraphics[height = 0.85\textheight]{slides/unit-3/chad-inspection.jpg}
\end{center}

## Bush v. Gore

. . .

- On December 8, 2000, the Florida Supreme Court ordered a recount of all contested ballots in Miami-Dade County and all "under-vote" ballots in Florida
- The Court held oral arguments on the 11th and issued the opinion on the 12th
- 7-2, the Court held that FL Supreme Court's standardless recount mandate violated the Equal Protection Clause
- 5-4, the Court held that there was no time to implement a permissible recount method

## Bush v. Gore: Criticisms: Others should decide

\begin{center}
\includegraphics[height = 0.5\textheight]{slides/unit-3/court-protest.jpg}
\end{center}

## Bush v. Gore: Criticisms: Politicians in robes

\begin{center}
\includegraphics[height = 0.5\textheight]{slides/unit-3/court-cartoon.jpg}
\end{center}

## Bush v. Gore: Criticisms: "Good for one day only"

\begin{block}{Bush v. Gore}
Our consideration is limited to the present circumstances....
\end{block}

## Bush v. Gore: Criticisms: "Good for one day only"

\begin{figure}
\centering
\includegraphics[width = 0.9\textwidth]{slides/unit-3/npr.png}
\end{figure}

## Bush v. Gore: Criticisms: "Good for one day only"

\begin{block}{From the transcript}
HANSEN: This is an interesting concept, the idea that the case can't be used as precedent. Isn't that one of the foundations of our legal system and it even has a Latin term? \vspace{0.5\baselineskip}

Prof. WESSON: Stari decisis, that's right. And almost everyone who's thought about what we call the rule of law, which is something we're really proud of in this country, agrees that a court's obligation to follow its own precedents - or at least to acknowledge them and explain why it's chosen not to follow them - is one of the most important features of the rule of law. Now, that's not to say that it's never justified for a court to overrule its own precedents.... \vspace{0.5\baselineskip}

So it's that contrary aspect of Bush vs. Gore, the Court's announcement that it was not going to treat its own work as precedential, that shocked and dismayed a lot of Court observers. 
\end{block}

## Bush v. Gore: Criticisms: "Good for one day only"

\includegraphics[width = 0.9\textwidth]{slides/unit-3/not-named.png}

## Controversy Over the Electoral College

. . .

\begin{center}
\includegraphics[height = 0.85\textheight]{slides/unit-3/trump-headline.png}
\end{center}

## Controversy Over the Electoral College

```{r kable, echo=FALSE}
table_df = data.frame(Election = c(1824, 1876, 1888, 2000, 2016),
                      President = c("John Quincy Adams",
                                    "Rutherford B. Hayes",
                                    "Benjamin Harrison",
                                    "George W. Bush",
                                    "Donald J. Trump"),
                      `Electoral College` = c(32.18, 50.14, 58.1, 50.37, 55.5),
                      `Popular Vote` = c(30.92, 47.92, 47.8, 47.87, 46.09),
                      `Popular Margin` = c(-10.44, -3.02, -0.79, -0.51, -2.1))
table_heads = c("Election", "President", "EC (%)", "Pop. Vote (%)",
                "Margin")
library(knitr)
kable(table_df, col.names = table_heads)
```

## Controversy Over the Electoral College

\begin{center}
\includegraphics[height = 0.85\textheight]{slides/unit-3/ec-critique.png}
\end{center}

## Controversy Over the Electoral College

\begin{center}
\includegraphics[height = 0.4\textheight]{slides/unit-3/trump-ec-tweet1.png}
\end{center}

## Controversy Over the Electoral College

\begin{center}
\includegraphics[height = 0.5\textheight]{slides/unit-3/trump-ec-tweet2.png}
\end{center}

## Controversy Over the Electoral College: Next?

. . .

- Who does the Electoral College disproportionately favor?
- National Popular Vote Interstate Compact (NPVIC)
- So far, Donald Trump is on track to win the popular vote for 2024
  + Some suggest coalitions supporting/opposing EC may shift
  + Stay tuned...

<!-- ## Foreign Policy -->

<!-- . . . -->

<!-- -  -->

## The Bureaucracy

. . .

- The bureaucracy
  + is a tool for presidential power, and
  + often considered necessary in modern times (*Mistretta v. U.S.*)
- Cracks have formed in the "intelligible principle" doctrine
  + *Gundy v. U.S.*
- The Court has now also eliminated "*Chevron* deference"
  + Bureaucrats sometimes have to interpret their enabling legislation
  + In *Chevron v. NRDC* the Court held judges defer to them if
    * the statute has an ambiguity, and
    * the interpretation is reasonable
  + In *Loper Bright Enterprises v. Raimondo*, the Court overruled *Chevron*
    * A "court shall decide all relevant questions of law" 5 U.S.C. \S 706
    * *Chevron* falsely assumed Congress *intends* to delegate interpretative authority
    * Further, *Chevron* was often applied inconsistently
