---
title: "Studying Judicial Decision Making"
author: "JBrandon Duck-Mayr"
date: "`r quack::american_date_format(Sys.Date())`"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}[1]{\setlength{\itemsep}{#1}}
    - \newcommand{\setskip}[1]{\setlength{\parskip}{#1}}
    - \renewcommand{\tightlist}{\setsep{3pt}\setskip{3pt}}
    - \definecolor{firebrick3}{RGB}{205,38,38}
    - \definecolor{midgreen}{RGB}{69,139,0}
    - \definecolor{dodgerblue4}{RGB}{16,78,139}
    - \usepackage{tikz}
    - \usetikzlibrary{arrows.meta}
    - \usepackage{wrapfig}
output:
    quack::presentation:
        toc: false
        incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
quack::attach_common_packages()
```

## Recap

. . .

- The judicial power is the power to "say what the law is"
- The Constitution creates the Supreme Court and the office of Chief Justice
  + Congress sets the structure of the lower courts
    * Created a three tier system
    * Made appellate courts "collegial" (more than one judge)
  + Congress controls the number of seats & (in part) the Court's jurisdiction
- The Court has used the judicial power to make big changes
- Now we turn to the factors affecting these judges' decisions

## Models of Judicial Decision Making

. . .

What are the factors that influences' judges decisions?

- **Legal model**: judges apply legal principles, precedent, and techniques of statutory and constitutional interpretation to the facts of cases to make decisions
- **Attitudinal model**: judges react purely according to their public policy preferences
- **Strategic model**: attitudinal model + strategic considerations like preserving public support to maintain institutional strength to achieve long-term goals

. . .

::: {.block}
### James L. Gibson, quoted in Baum (1997)
Judges' decisions are a function of what they prefer to do, tempered by what they think they ought to do, but constrained by what they perceive is feasible to do.
:::

## Modeling Judicial Decision Making: Techniques

- How do we apply these ideas to analyze judges' behavior in various contexts?
- How do we determine what judges' political preferences are?
- How do we measure the influence of public opinion or other political constraints?
- How do we determine the effect of precedent and other legal factors?

## Measuring Preferences

. . .

```{r agreement, out.height = "90%", fig.height = 5, fig.width = 7, fig.align = "center"}
scdb  = read_csv("SCDB_2023_01_justiceCentered_Citation.csv")
votes = scdb %>% 
    filter(term == 2022) %>% 
    select(caseId, justiceName, majority) %>% 
    pivot_wider(names_from = "caseId", values_from = "majority") %>% 
    tibble::column_to_rownames("justiceName") %>% 
    as.matrix()
unanimous = apply(votes, 2, function(x) length(unique(na.omit(x))) == 1)
votes = votes[ , !unanimous]
justices = gsub("^[A-Z]+([A-Z][a-z]+)$", "\\1", rownames(votes))
rownames(votes) = justices
levs = c(
    "Alito", "Thomas", "Gorsuch", "Barrett", "Roberts",
    "Kavanaugh", "Jackson", "Kagan", "Sotomayor"
)
justices = factor(justices, levels = levs)
dat = expand.grid(x = justices, y = justices)
dat$agreement = 100 * sapply(1:nrow(dat), function(i) {
    row1 = as.character(dat$x[i])
    row2 = as.character(dat$y[i])
    return(mean(votes[row1, ] == votes[row2, ], na.rm = TRUE))
})
ttl = "Agreement in non-unanimous cases (2022 term)"
cap = "Data from the Supreme Court Database"
map = aes(x = x, y = y, fill = agreement)
ggplot(data = dat, mapping = map) +
    geom_tile() +
    labs(title = ttl, caption = cap) +
    scale_fill_gradient2(
        high = "#005f86",
        low = "#bf5700",
        midpoint = 50,
        name = ""
    ) +
    theme(
        axis.title = element_blank(),
        plot.caption = element_text(hjust = 0, color = "#5f5f5f")
    )
```

## Measuring Preferences: "Ideal points"

```{r mq-scores, out.width = "90%", fig.height = 2, fig.width = 8, fig.align = "center"}
mq = read_csv("justices.csv")
dat = mq %>%
    filter(term == 2022) %>% 
    mutate(justiceName = gsub("^[A-Z]+([A-Z][a-z]+)$", "\\1", justiceName)) %>% 
    arrange(post_mn) %>% 
    mutate(
        y = rep(c(-0.5, 0.5), length.out = 9),
        h = c(0, 1, 0, 1, 0.75, 0.25, 0.25, 0.5, 0.5)
    )
ggplot(data = dat, mapping = aes(x = post_mn, y = y)) +
    geom_hline(yintercept = 0, linewidth = 1) +
    geom_point(aes(y = 0), size = 3) +
    geom_text(aes(label = justiceName, hjust = h)) +
    ylim(-1, 1) +
    ggtitle("Martin-Quinn scores for 2022 term") +
    theme_void() +
    theme(plot.caption = element_text(hjust = 0, color = "#5f5f5f"))
```

## How do preferences turn into policy?

\begin{figure}
\centering
\begin{tikzpicture}
    \node (start) at (-3, 0) {};
    \node (end) at (3, 0) {};
    \draw[->, thick] (start)--(end);
    \node[right] at (end.west) {More intrusive};
    \node[below] at (-2, -1) {Judge 1};
    \node[below] at ( 0, -1) {Judge 2};
    \node[below] at ( 2, -1) {Judge 3};
    \draw (-2, -0.75)--(-2, 0.75);
    \draw ( 0, -0.75)--( 0, 0.75);
    \draw ( 2, -0.75)--( 2, 0.75);
    \only<2>{
    \draw[fill = black] (-2.5, 0) circle (0.1);
    }
    \only<3-10>{
    \draw[fill = Blue, draw = Blue] (-2.5, 0) circle (0.1);
    }
    \only<4>{
    \draw[fill = black] (-1.5, 0) circle (0.1);
    }
    \only<5-10>{
    \draw[fill = Blue, draw = Blue] (-1.5, 0) circle (0.1);
    }
    \only<6>{
    \draw[fill = black] (1, 0) circle (0.1);
    }
    \only<7-10>{
    \draw[fill = BurntOrange, draw = BurntOrange] (1, 0) circle (0.1);
    }
    \only<8>{
    \draw[fill = black] (2.5, 0) circle (0.1);
    }
    \only<9-10>{
    \draw[fill = BurntOrange, draw = BurntOrange] (2.5, 0) circle (0.1);
    }
    \only<10>{
    \draw[fill = Blue, draw = Blue] (-1.00, 0) circle (0.1);
    \draw[fill = Blue, draw = Blue] (-0.50, 0) circle (0.1);
    \draw[fill = Blue, draw = Blue] (-0.75, 0) circle (0.1);
    \draw[fill = BurntOrange, draw = BurntOrange] ( 0.50, 0) circle (0.1);
    \draw[fill = BurntOrange, draw = BurntOrange] ( 1.50, 0) circle (0.1);
    }
    \only<11>{
    \draw[fill = Blue, draw = Blue] (-3, 0.1) rectangle (0, -0.1);
    \draw[fill = BurntOrange, draw = BurntOrange] (0, 0.1) rectangle (2.75, -0.1);
    \draw (-2, -0.75)--(-2, 0.75);
    \draw ( 0, -0.75)--( 0, 0.75);
    \draw ( 2, -0.75)--( 2, 0.75);
    }
\end{tikzpicture}
\end{figure}

## A wrinkle: There's not just one kind of case!

. . .

[Clark and Lauderdale 2012](https://doi.org/10.1017/S0003055412000469):

\only<2>{
\begin{figure}
\tiny
\centering
\includegraphics[height = 0.8\textheight]{many-medians}
\end{figure}
}

\only<3>{
\begin{figure}
\centering
\includegraphics[width = 0.95\textwidth]{many-medians2}
\end{figure}
}

## Plus... are issues themselves multidimensional?!

\begin{figure}
\centering
\begin{tikzpicture}
\only<2>{
    \node (start) at (-3, 0) {};
    \node (end) at (3, 0) {};
    \draw[->, thick] (start)--(end);
    \node[right] at (end.west) {More intrusive};
}
\only<4>{
    \fill[fill = Blue] (0, 0)--(2.5, 0)--(0, 2.5)--cycle;
    \fill[fill = BurntOrange] (2.5, 0)--(0, 2.5)--(0, 4.75)--(4.75, 0)--cycle;
}
\onslide<3->{
    \draw[->, thick] (0, 0)--(5, 0);
    \draw[->, thick] (0, 0)--(0, 5);
    \node[below] at (2.5, -0.1) {Intrusiveness};
    \node[left, rotate=90] at (-0.3, 4.5) {Evidence (decreasing)};
}
\end{tikzpicture}
\end{figure}

## Wait, *how* does the Court make policy again??

. . .

- The Court *does* make policy in the settings of cases...
- But we don't **just** see the outcomes
- They also give us opinions!
  + These opinions explain **why** the case got the outcome it did
  + Some write this off as post-hoc justification
  + But don't be so hasty! Lower courts & others pay attention
- So we can learn about policy from opinion content
  + Dealing with the text for measurement is difficult, but ongoing
- Theoretical work in this space gives us good reason to think the median *of the majority coalition* rather than the median of the Court has the most influence on opinion content ([Carrubba et al 2012](https://doi.org/10.1111/j.1540-5907.2011.00557.x))

## Be careful---These models are abstractions

::: {.block}
### George E.P. Box
All models are wrong; some models are useful.
:::

## And don't forget there are other factors...

. . .

- External constraints
  + Public opinion
  + Other branches
- Internal constraints
  + \textbf<8>{\textcolor<8>{BurntOrange}{Legal factors}}
  + Hierarchy and collegiality

\invisible<1-2, 4-8>{
\begin{block}{Justice Owen Roberts (Interview by Merlo J. Pusey)}
It is difficult to say what makes a judge decide as he does.
Public outcry against an opinion is bound to have some effect on a man's thinking when it is a question of degree---of how far we can go.
\end{block}
}

## One final note about ideal points...

. . .

- Ideology measures are (most) often derived from judges' votes
- Sometimes we care about how ideology affects judges' votes
- Can we use such measures as explanatory variables?
- [Martin and Quinn (2005)](https://mqscores.lsa.umich.edu/media/resnote.pdf)

## "Political jurisprudence"

::: {.block}
### C. Herman Pritchett
\textcolor<2-3>{LightGray}{[P]olitical scientists who have done so much to put the "political" in "political jurisprudence" need to emphasize that it is still "jurisprudence."}
\textcolor<1,3>{LightGray}{It is judging in a political context, but it is still judging; and judging is something different from legislating...
Judges make choices, but they are not the "free" choices of Congressmen...}
\textcolor<1-2>{LightGray}{Any accurate analysis of judicial behavior must have as a major purpose a full clarification of the unique limiting conditions under which judicial policy making proceeds.}
:::

## What kinds of "limiting conditions"?

- Procedural
  + "The judicial power shall extend to... cases... [and] controversies..." US Constitution, Article III \S 2
  + Statutory limitations
  + Aversion to *sua sponte* issue consideration
- Substantive
  + "Easy" vs. "hard" cases
  + Power of the text
  + *Stare decisis*

## **Why** would judges care?

## Cases and Controversies

::: {.block}
### US Constitution, Article III, \S 2
The judicial power shall extend to all cases... arising under this Constitution, the laws of the United States, and treaties made... under their authority;--to all cases affecting [various parties];--to controversies [between various parties]...
:::

. . .

- So federal judges only have the power to decide certain things
- Some categories of things judges will therefore avoid:
  + Advisory opinions
  + Cases without a party with standing
  + Cases that are moot or are not yet ripe
  + "Political questions"

## *Stare decisis*

\only<2-6>{
\begin{itemize}
\tightlist
\item<2-> A constraining role of law is somewhat controversial in political science
\item<3-> Ideological variables are highly predictive of votes
\item<4-> And some older studies show little respect for \textit{stare decisis}
\item<5-> But we see intra-Court discussions about precedent
\item<6-> And recent scholarship has shown renewed interest in the issue
\end{itemize}
}

\only<7>{
\begin{figure}
\begin{columns}
\column{0.3\linewidth}
\href{https://doi.org/10.1017/S0003055409990049}{Bartels (2009)} Figure 1: Constraining Capacity of Legal Rules: (A) Legal Presumptions Model, and (B) Rights-Protectiveness Model
\column{0.65\linewidth}
\includegraphics[height = 0.8\textheight]{bartels1}
\end{columns}
\end{figure}
}

\only<8>{
\begin{figure}
\begin{columns}
\column{0.3\linewidth}
\href{https://doi.org/10.1017/S0003055409990049}{Bartels (2009)} Figure 4: Predicted Magnitudes of Ideological Voting for Each Free Expression Category Before and After the Grayned Doctrine: (A) Post-Grayned and (B) Pre-Grayned
\column{0.65\linewidth}
\includegraphics[height = 0.8\textheight]{bartels2}
\end{columns}
\end{figure}
}

\only<9>{
Difficulty:

\begin{figure}[htb]
    \centering
    \begin{tikzpicture}[thick,->]
        \node (X1) at ( 0,  0) {\( \mathbf{x}_1 \)};
        \node (L1) at ( 2,  2) {\( \lambda_1 \)};
        \node (P1) at ( 2, -2) {\( \phi_1 \)};
        \node (Y1) at ( 4,  0) {\( y_1 \)};
        \path (X1) edge (L1);
        \path (X1) edge (P1);
        \path (L1) edge (Y1);
        \path (P1) edge (Y1);
        \node (X2) at ( 6,  0) {\( \mathbf{x}_2 \)};
        \node (L2) at ( 8,  2) {\( \lambda_2 \)};
        \node (P2) at ( 8, -2) {\( \phi_2 \)};
        \node (Y2) at (10,  0) {\( y_2 \)};
        \path (X2) edge (L2);
        \path (X2) edge (P2);
        \path (L2) edge (Y2);
        \path (P2) edge (Y2);
        \path (P1) edge (P2);
        \path (L1) edge (L2);
        \path[dashed] (Y1) edge (L2);
    \end{tikzpicture}
    \caption{
        Directed acyclic graph for the influence on a justice's decisions.
        \( \mathbf{x}_i \) denotes the facts of case \( i \),
        \( y_i \) denotes the justice's decision in case \( i \),
        \( \lambda_i \) denotes the predicted legal outcome in case \( i \),
        and \( \phi_i \) represents the justice's policy preferences
        in case \( i \).
    }
    \label{fig:dag}
\end{figure}
}

## Legal constraints or political opportunities?
