---
title: "Research on American Courts"
author: "JBrandon Duck-Mayr"
date: "`r quack::american_date_format(Sys.Date())`"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}[1]{\setlength{\itemsep}{#1}}
    - \newcommand{\setskip}[1]{\setlength{\parskip}{#1}}
    - \renewcommand{\tightlist}{\setsep{3pt}\setskip{3pt}}
    - \definecolor{FireBrick3}{RGB}{205,38,38}
    - \definecolor{midgreen}{RGB}{69,139,0}
    - \definecolor{dodgerblue4}{RGB}{16,78,139}
    - \usepackage{tikz}
    - \usetikzlibrary{arrows.meta}
    - \usepackage{wrapfig}
    - \newcommand{\tcbf}[2]{\textcolor{#1}{\textbf{#2}}}
output:
    quack::presentation:
        toc: false
        incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
quack::attach_common_packages()
theme_set(theme_bw())
theme_update(axis.title = element_blank())
theme_update(plot.caption = element_text(color = "#5f5f5f", hjust = 0))
```

## Judicial Politics Research

. . .

- Judicial politics research has evolved quite a bit
  + The initial focus was completely on "public law"
  + Then political scientists took some baby steps into empirical research about a century ago
  + Quantitative methods became mainstream in judicial politics around 50 years ago
  + Formal theoretic work became more common in the 90s
- Covers a broad range of topics
  + \textbf<14>{\textcolor<14>{BurntOrange}{Selecting judges}}
  + \textbf<14>{\textcolor<14>{BurntOrange}{Agenda Setting}}
  + Measuring ideology
  + Assessing ideological effects in various settings
  + \textbf<14>{\textcolor<14>{BurntOrange}{Effects of institutions like hierarchy and collegiality}}
  + and more

## Selecting Judges

## Selecting judges: What do we want?

## Selecting federal judges: How does it work?

. . .

::: {.block}
### US Constitution Article II, Section 2
...\textcolor<5>{BurntOrange}{\textbf<5>{[the president] shall nominate}}, \textcolor<6>{BurntOrange}{\textbf<6>{and by and with the advice and consent of the Senate, shall appoint}}... \textcolor<3>{BurntOrange}{\textbf<3>{judges of the Supreme Court}}, \textcolor<4>{BurntOrange}{\textbf<4>{and all other officers of the United States, whose appointments are not herein otherwise provided for, and which shall be established by law}}...

&nbsp;

\textcolor<7>{BurntOrange}{\textbf<7>{The President shall have power to fill up all vacancies that may happen during the recess of the Senate}}, \textcolor<8>{BurntOrange}{\textbf<8>{by granting commissions which shall expire at the end of their next session.}}
:::

## Selecting federal judges: How does it work?

- So all federal judges are nominated by the president
- If the Senate is in recess, they can unilaterally make a "recess appointment"
  + But recess appointments expire at the end of the session
- If not, the Senate must confirm the nomination
  + This is by majority vote\invisible<5>{, but}
  + various Senate procedures can prevent a vote

\onslide<7>{
\begin{block}{US Constitution Article I, Section 5}
...Each House may determine the rules of its proceedings...
\end{block}
}

## Picking candidates

- White House, DOJ, & other executive staff help find & vet candidates
- For lower courts, Senators & party leaders may propose candidates
- Interest groups also play a role in finding & vetting candidates

## Senate Confirmation Procedure

- Upon nomination, it goes to the Senate Judiciary Committee
- For District Court Judges, they consult home state senators ("blue slips") 
- The next step is a Judiciary Committee hearing
  + Often pro forma for lower courts, especially District Courts
  + Much bigger deal for Supreme Court nominees
- Then committee votes & sends recommendation to the full Senate
- Then the Senate votes on the nominee (typically confirms)
  + Though this has gotten more contentious...

. . .

## Confirmation Votes Over Time

```{r scotus-nominations, out.width = "90%", fig.height = 2.67, fig.width = 4}
dat = read_csv("scotus-nominations.csv") %>% 
    mutate(Date = as.Date(ResultDate, format = "%b %d, %Y")) %>% 
    filter(!is.na(Vote)) %>% 
    mutate(Vote = ifelse(Vote == "V", "100-0", Vote)) %>% 
    separate(col = "Vote", sep = "-", into = c("Yea", "Nay")) %>% 
    mutate(across(Yea:Nay, as.numeric)) %>% 
    mutate(Percent = 100 * Yea / (Yea + Nay)) %>% 
    mutate(Result = ifelse(Percent > 50, "Confirmed", "Not"))
ttl = "Percent of 'Yes' Votes on SCOTUS Nominees"
cap = "Data from Senate.Gov"
pal = c("Confirmed" = "#005f86", "Not" = "#bf5700")
pts = c("Confirmed" = 19, "Not" = 17)
map = aes(x = Date, y = Percent, color = Result, shape = Result)
ggplot(data = dat, mapping = map) +
    geom_point(size = 2, alpha = 3/4) +
    scale_color_manual(values = pal, name = "") +
    scale_shape_manual(values = pts, name = "") +
    labs(title = ttl, caption = cap) +
    ylim(0, 100) +
    theme(legend.margin = margin())
```

## Confirmation Votes: Bork!

. . .

\only<2>{
\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{Bork}
\end{figure}
}

\only<3>{
\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{Nixon}
\end{figure}
}

\only<4>{
\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{watergate-complex}
\end{figure}
}

\only<5>{
\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{saturday-night-massacre}
\end{figure}
}

\only<6-11>{
Red flags on Bork:
\begin{itemize}
\tightlist
\item<6-11> Role in the ``Saturday Night Massacre''
\item<7-11> Defended Virginia poll tax as ``a very small tax [that] was not discriminatory''
\item<8-11> Described the Civil Rights Act of 1964 as state coercion
\item<9-11> Opposed right to privacy
\item<10-11> In other words, \textcolor<11>{BurntOrange}{\textbf<11>{some}} viewed him as an extremist with a taint of scandal
\end{itemize}
}

\only<12>{
Confirmation vote for Bork:
\vspace*{\baselineskip}
\begin{table}
\begin{tabular}{lcc}
                & Yea       & Nay       \\
Republicans     & \tcbf{FireBrick3}{41} & \tcbf{FireBrick3}{6} \\
Democrats       & \tcbf{DodgerBlue4}{2} & \tcbf{DodgerBlue4}{52}
\end{tabular}
\end{table}
}

## Confirmation Votes: Bork!

```{r scotus-nominations2, out.width = "90%", fig.height = 2.67, fig.width = 4}
dat = read_csv("scotus-nominations.csv") %>% 
    mutate(Date = as.Date(ResultDate, format = "%b %d, %Y")) %>% 
    filter(!is.na(Vote)) %>% 
    mutate(Vote = ifelse(Vote == "V", "100-0", Vote)) %>% 
    separate(col = "Vote", sep = "-", into = c("Yea", "Nay")) %>% 
    mutate(across(Yea:Nay, as.numeric)) %>% 
    mutate(Percent = 100 * Yea / (Yea + Nay)) %>% 
    mutate(Result = ifelse(Percent > 50, "Confirmed", "Not"))
ttl = "Percent of 'Yes' Votes on SCOTUS Nominees"
cap = "Data from Senate.Gov"
pal = c("Confirmed" = "#005f86", "Not" = "#bf5700")
pts = c("Confirmed" = 19, "Not" = 17)
map = aes(x = Date, y = Percent, color = Result, shape = Result)
ggplot(data = dat, mapping = map) +
    geom_point(size = 2, alpha = 3/4) +
    geom_vline(
        xintercept = as.Date("1987-10-23"),
        linetype = "dashed", linewidth = 1
    ) +
    scale_color_manual(values = pal, name = "") +
    scale_shape_manual(values = pts, name = "") +
    labs(title = ttl, caption = cap) +
    ylim(0, 100) +
    theme(legend.margin = margin())
```


## Quantitative and theoretical research

. . .

- [Bird and McGee 2023](https://doi.org/10.1177/1532673X221109534)\only<2>{\small, "Going Nuclear: Federalist Society Affiliated Judicial Nominees’ Prospects and a New Era of Confirmation Politics"}
- [Cameron et al 2013](https://doi.org/10.1017/S0022381613000017)\only<3>{\small, "Voting for Justices: Change and Continuity in Confirmation Voting 1937–2010"}
- [Epstein et al 2006](https://doi.org/10.1111/j.1468-2508.2006.00407.x)\only<4>{\small, "The Changing Dynamics of Senate Voting on Supreme Court Nominees"}
- [Cameron and Kastellec 2016](https://doi.org/10.1017/S0003055416000496)\only<5>{\small, "Are Supreme Court Nominations a Move-the-Median Game?"}
- [Caldeira and Wright 1998](https://doi.org/10.2307/2991768)\only<6>{\small, "Lobbying for Justice: Organized Interests Supreme Court Nominations, and United States Senate"}
- [Cameron and Park 2011]( https://doi.org/10.1111/j.1741-5705.2011.03881.x)\only<7>{\small, "Going Public When Opinion Is Contested: Evidence from Presidents' Campaigns for Supreme Court Nominees"}

\invisible<1-8>{placeholder}

## Who are these federal judges?

\begin{itemize}
\tightlist
\item<2-> What are the official requirements to be a federal judge? \invisible<2>{\textcolor<3->{BurntOrange}{\textbf<3->{None}}}
\item<4-> What about unofficial requirements?
    \begin{itemize}
    \tightlist
    \item<5-> Ideology
    \item<6-> Pedigree
    \item<7-> Political activity
    \item<8-> Others?
    \end{itemize}
\end{itemize}

## Who are these federal judges?

::: {.block}
### Oliver Wendell Holmes, Jr., *The Common Law* (1881)
The life of the law has not been logic; it has been experience.
:::

. . .

Federal judges are disproportionately

. . .

- wealthy
- from "elite" schools
- from "elite" backgrounds
- white, male, etc
- prosecutors rather than defenders

## Who are these federal judges?

```{r judge-demos, out.width = "90%", fig.height = 2.67, fig.width = 4, fig.align = "center"}
demos = read_csv("judge-demographics.csv")
years_active = read_csv("judge-years-active.csv")
dat = demos %>% left_join(years_active, by = "jid") %>% 
    filter(!is.na(confirmed)) %>% 
    mutate(year = purrr::map2(confirmed, terminated, seq)) %>% 
    unnest(year) %>% 
    group_by(year) %>% 
    summarise(
        percent_female = 100 * sum(Gender != "Male") / n(),
        percent_nonwhite = 100 * sum(Race != "White") / n(),
        percent_nonWM = 100 * sum(Race != "White" | Gender != "Male") / n()
    ) %>%
    pivot_longer(
        cols = starts_with("percent"),
        names_to = "group",
        names_prefix = "percent_",
        values_to = "percent"
    ) %>% 
    filter(year >= 1948)

map = aes(x = year, y = percent, color = group, linetype = group)
pal = c(female = "#0072b2", nonwhite = "#d55e00", nonWM = "black")
lty = c(female = "solid", nonwhite = "dashed", nonWM = "dotted")
nms = c("Female", "Non-white", "Not a white man")
ttl = "% of federal judges that are not white/men"
cap = "Data from the Federal Judicial Center"
ggplot(data = dat, mapping = map) +
    geom_hline(yintercept = 69.0, color = "#00000080") +
    geom_hline(yintercept = 50.5, color = "#0072b280") +
    geom_hline(yintercept = 38.4, color = "#d55e0080") +
    geom_line(linewidth = 1) +
    scale_color_manual(values = pal, labels = nms, name = "") +
    scale_linetype_manual(values = lty, labels = nms, name = "") +
    scale_y_continuous(limits = c(0, 100)) +
    labs(title = ttl, caption = cap) +
    theme(
        axis.title = element_blank(),
        plot.caption = element_text(hjust = 0, color = "#5f5f5f"),
        legend.box.margin = margin(),
        legend.box.spacing = unit(0, "in"),
        legend.key.width = unit(0.4, "in"),
        legend.position = "bottom",
        legend.direction = "horizontal"
    )
```

## Who are these federal judges?

```{r judge-demos2, out.width = "90%", fig.height = 2.67, fig.width = 4, fig.align = "center"}
backgrounds = read_csv("judge-backgrounds.csv")
years_active = read_csv("judge-years-active.csv")
top_law_schools = c("Harvard", "Yale", "Stanford", "Chicago", "Columbia")
top_undergrads = c("Harvard", "Princeton", "Yale", "Stanford")
levs = c("prosecutor", "defender", "elite_law_school", "elite_undergrad")
match_any = function(text, targets) grepl(paste(targets, collapse = "|"), text)
dat = backgrounds %>% left_join(years_active, by = "jid") %>% 
    filter(terminated >= 2022) %>% 
    mutate(elite_law_school = match_any(law_school, top_law_schools)) %>%
    mutate(elite_undergrad  = match_any(undergrad,  top_undergrads)) %>%
    summarise(
        percent_elite_undergrad  = 100 * sum(elite_undergrad) / n(),
        percent_elite_law_school = 100 * sum(elite_law_school) / n(),
        percent_prosecutor = 100 * sum(prosecutor) / n(),
        percent_defender = 100 * sum(defender) / n()
    ) %>%
    pivot_longer(
        cols = starts_with("percent"),
        names_to = "group",
        names_prefix = "percent_",
        values_to = "percent"
    ) %>% 
    mutate(type = ifelse(grepl("elite", group), "School", "Occupation")) %>% 
    mutate(group = factor(group, levels = levs))

map = aes(x = group, y = percent, fill = group)
pal = c(
    elite_law_school = "#56b4e9", elite_undergrad = "#009e73",
    prosecutor = "#d55e00", defender = "#0072b2"
)
nms = c("Elite\nlaw school", "Elite\nundergrad", "Prosecutor", "Defender")
ttl = "Backgrounds of current federal judiciary"
cap = "Data from the Federal Judicial Center"
x_labeller = function(x) tools::toTitleCase(sub("_", " ", sub("_", "\n", x)))
ggplot(data = dat, mapping = map) +
    facet_wrap(~type, scales = "free_x") +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = pal, labels = nms, name = "") +
    scale_y_continuous(limits = c(0, 50), labels = ~ sprintf("%d%%", .x)) +
    scale_x_discrete(labels = x_labeller) +
    labs(title = ttl, caption = cap) +
    theme(
        axis.title = element_blank(),
        plot.caption = element_text(hjust = 0, color = "#5f5f5f"),
        legend.position = "none"
    )
```

## Agenda Setting

The importance of agenda setting

. . .

- Answers to questions come only after deciding what questions *to* answer
- Through agenda control, "some issues are organized into politics while others are organized out" (Schattschneider 1960)

. . .

Common topics in judicial agenda setting

. . .

- Why Supreme Court hears particular cases
  + Role of organized interests
  + Role of legal factors
  + Role of ideology
  + And more
- How the composition of the Court's docket changes over time
- Top-down versus bottom-up forces

## Supreme Court Agenda Setting Process: Recap

::: {.nonincremental}
- Supreme Court has discretion over whether to hear almost all cases sent to it since the Supreme Court Case Selections Act of 1988
- So first a party *asks* the Court to hear their case
- Sometimes outside parties file *amici* briefs asking the Court to take the case (or not)
- The petition is distributed to the "cert pool"
- About 30% of petitions then go on the "discuss list"; all others summarily denied
- If $\geq 4$ justices vote to hear the case, they hear it ("Rule of Four")
:::

## Supreme Court Agenda Setting: Shrinking Docket

```{r full-hearing, out.width = "90%", fig.height = 2.67, fig.width = 4, fig.align = "center"}
dat = read_csv("scotus-caseload.csv") %>%
    filter(grepl("granted", category))
ggplot(data = dat, mapping = aes(x = term, y = total)) +
    geom_vline(xintercept = 1987, linetype = "dashed", alpha = 1/2) +
    geom_line(linewidth = 1, color = "#005f86") +
    scale_y_continuous(limits = c(0, NA)) +
    ggtitle("# of cases given full hearing at SCOTUS")
```

## Supreme Court Agenda Setting: Shrinking Docket

```{r grant-rate, out.width = "90%", fig.height = 2.67, fig.width = 4, fig.align = "center"}
dat = read_csv("scotus-caseload.csv") %>%
    filter(grepl("granted|Docketed", category)) %>% 
    select(term, category, total) %>% 
    mutate(category = ifelse(grepl("grant", category), "heard", "filed")) %>%
    pivot_wider(names_from = "category", values_from = "total") %>% 
    mutate(grant_rate = heard / filed)
ggplot(data = dat, mapping = aes(x = term, y = grant_rate)) +
    geom_vline(xintercept = 1987, linetype = "dashed", alpha = 1/2) +
    geom_line(linewidth = 1, color = "#005f86") +
    scale_y_continuous(limits = c(0, NA), labels = scales::percent) +
    ggtitle("% of filings given full hearing")
```

## Role of organized interests

. . .

- Informational role as *amici* for cert (Caldeira and Wright 1988)
- Role as litigants

## Role of legal factors

. . .

- Things like "circuit splits" reliable predictors of cert
- Can dampen ideological forces or provide opportunities (Black and Owens 2009)

. . .

::: {.block}
### Perry (1991), 269; emphasis supplied
I would say that [cert votes] are sometimes tentative votes on the merits. Now I would say that there are certain cases that I would vote for, for example, if there was a clear split in circuits, *I would vote for cert. without even looking at the merits*. But there are other cases I would have more of a notion of what the merits were.
:::

## Compositional studies

. . .

- We may care less about whether SCOTUS takes a specific case than relative time spent on economics versus civil liberties
- A number of studies take this approach (e.g. Baird 2004; Peters 2007; Pacelle 1991; Flemming, Bohte, and Wood 1997; Flemming, Wood, and Bohte 1999; Hurwitz 2006)
- For example, Baird (2004) finds that an increase in the number of cases the Court hears in a particular issue area influences the number of cases they hear in that area in the future

## Bottom-up forces

\only<2-4>{
\begin{itemize}
\tightlist
\item<2-4> How much can the lower courts affect the Supreme Court's agenda?
\item<3-4> Hurwitz (2006) finds when appellate courts hear more economic cases, afterward so does SCOTUS, while when SCOTUS hears more civil liberties cases, afterward so do the appellate courts
\item<4> Duck-Mayr, Hansford, and Spriggs (2021) find directions in both directions for a different conception of agenda: attention paid to precedents
\end{itemize}
}

\only<5>{
Supreme Court precedents from 1946 to 1986 with each dynamic:

\vspace*{\baselineskip}

\begin{table}
\centering
\begin{tabular}{lccc}
\toprule
                    & SC Cites$_t$  & AC Cites$_t$  & DC Cites$_t$  \\
\midrule
SC Cites$_{t-1}$    & 652           & 422           & 458           \\
AC Cites$_{t-1}$    & 1919          & 1317          & 1017          \\
DC Cites$_{t-1}$    & 2185          & 1111          & 1696          \\
\bottomrule
\end{tabular}
\end{table}
}

\only<6>{
\begin{figure}
    \centering
    \begin{tikzpicture}[->]
        \node (SC) at ( 0.0,  0.0) {Supreme Court};
        \node (AC) at ( 4.8, -2.5) {Circuit Courts};
        \node (DC) at (-4.8, -5.0) {District Courts};
        \draw (SC) edge[bend left = 10] node[fill=white] {155} (AC); 
        \draw (SC) edge[bend left = 10] node[fill=white] {158} (DC); 
        \draw (AC) edge[bend left = 10] node[fill=white] {312} (SC); 
        \draw (AC) edge[bend left = 10] node[fill=white] {509} (DC); 
        \draw (DC) edge[bend left = 10] node[fill=white] {570} (SC); 
        \draw (DC) edge[bend left = 10] node[fill=white] {661} (AC); 
    \end{tikzpicture}
\end{figure}
}

\only<7>{
\begin{table}
\centering
\begin{tabular}{lc}
\toprule
& Time Series Granger Causes Both Other Series \\
\cmidrule{2-2}
SC Cites$_{t-1}$    & 15 \\
AC Cites$_{t-1}$    & 63 \\
DC Cites$_{t-1}$    & 97 \\
\midrule
& Time Series Granger Causes Both Other Series      \\
& \emph{and} No Other Series Granger Causes Another \\
\cmidrule{2-2}
SC Cites$_{t-1}$    & 12 \\
AC Cites$_{t-1}$    & 59 \\
DC Cites$_{t-1}$    & 84 \\
\midrule
& Only Lag Coefficients in One Direction Differ From 0 \\
\cmidrule{2-2}
Top-Down    & 30  \\
Bottom-Up   & 324 \\
\bottomrule
\end{tabular}
\end{table}
}

## The "Hierarchy of Justice"

```{=latex}
\begin{figure}
\centering
\begin{tikzpicture}
\node[draw = black, fill = white, rectangle, rounded corners] (SC) at (0,  3) {US Supreme Court};
\node[draw = black, fill = white, rectangle, rounded corners] (AC) at (0,  0) {US Courts of Appeals};
\node[draw = black, fill = white, rectangle, rounded corners] (DC) at (0, -3) {US District Courts};
\onslide<2>{
\draw[->, thick, shorten >=0.1cm] (DC) edge node[fill=white] {Must obey} (AC.south);
\draw[->, thick, shorten >=0.1cm] (AC) edge node[fill=white] {Must obey} (SC.south);
}
\end{tikzpicture}
\end{figure}
```

## Principal-agent problems

. . .

- Often an agent are charged with working on behalf of a principal
  + The appellate courts are the agents of the Supreme Court
  + The district courts are the agents of the Supreme Court and the appellate court above them
- However, the agent may have different preferences and the principal may not be able to monitor the agent perfectly
- So sometimes delegation results in "agency loss"
- In particular, sometimes lower courts don't obey higher courts even though they're supposed to
- Several studies have looked at how disobedient lower courts are, under what circumstances, etc

## What does it mean to be faithful to the principal?

```{=latex}
\begin{figure}
\centering
\begin{tikzpicture}
\node (P) at (0,  4) {\textit{Miranda v. Arizona}};
\node[draw = black, fill = white, rectangle, rounded corners] (SC) at (0,  3) {1966 Supreme Court};
\node[draw = black, fill = white, rectangle, rounded corners] (LC) at (0,  0) {1966 Lower Courts};
\draw[->, thick, shorten >=0.1cm] (LC) edge node[fill=white] {Must obey} (SC.south);
\onslide<2->{
\node[draw = black, fill = white, rectangle, rounded corners] (SC2) at (6,  3) {2023 Supreme Court};
\node[draw = black, fill = white, rectangle, rounded corners] (LC2) at (6,  0) {2023 Lower Courts};
\draw[->, thick, shorten >=0.1cm] (LC2) edge node[fill=white] {Obey this?} (SC.south east);
}
\onslide<3->{
\draw[->, thick, shorten >=0.1cm] (LC2) edge node[fill=white] {Or this?} (SC2.south);
}
\end{tikzpicture}
\end{figure}

\onslide<4>{
\vspace*{\baselineskip}

This is the issue brought up by Westerland et al (2010)
}
```

## *Some* agency loss

. . .

- Overall, lower courts are mostly responsive to the Supreme Court
- **But**, *some* "agency loss" occurs
- Past lower court defiance can make future lower court defiance easier (Westerland et al 2010)

## Informational functions of the hierarchy

. . .

- Clarification of issues as cases progress
- Lower court allies can help signal which cases need error correction ([Kastellec 2007](https://doi.org/10.1093/jleo/ewm026); [Beim, Hirsch, and Kastellec 2014](https://doi.org/10.1111/ajps.12108))
- Lower court allies can help justices understand implications of precedent ([Hansford, Spriggs, and Stenger 2013](https://doi.org/10.1017/S0022381613000686))

## Collegiality

. . .

- Well-known issues of collegial judgment aggregation
  + The "Doctrinal Paradox" ([Kornhauser 1992](https://www.jstor.org/stable/764863))
  + Extension of the paradox to rulemaking ([Landa and Lax 2009](https://doi.org/10.1017/S0022381609090811))
  + The paradox is exacerbated for mixed questions ([Duck-Mayr 2022](https://doi.org/10.1177/09516298211061159))
- But there are other interesting effects too: "panel effects"
  + The impact of circuit judges' ideology is muted by "counterjudges" ([Kastellec 2011](https://doi.org/10.1017/s0022381611000314))
  + This could be because they want to avoid review ([Kastellec 2011](https://doi.org/10.1017/s0022381611000314); [Beim, Hirsch, and Kastellec 2014](https://doi.org/10.1111/ajps.12108))
  + Or could be norms or workload considerations ([Epstein, Landes, and Posner 2011](https://www.hup.harvard.edu/catalog.php?isbn=9780674049895))
  + Or even due to deliberation ([Spitzer and Talley 2013](https://doi.org/10.1093/jleo/ewr013))
  + Identity of panel members affects opinion content ([Haire, Moyer, and Treier 2013](https://doi.org/10.1086/670724))
